---
title: "5: Introduction to machine learning - Exercise"
author: "Wilhelmine Bach, Dominik Kirschner, Fabian Fopp & Camille Albouy"
date: "22 8 2022"
output: html_document
html_document: default
---

### Intro 
Now that you have learned the basics of LM, GLM and unsupervised machine learning you should be able to solve this exercise. 

You will again work on the bird data set you know already from the tutorial. In addition, we will provide you with a number of additional species occurring in Switzerland. You will have land cover information (e.g. forest cover) for each grid cell as well as basic environmental data (e.g. temperature). 

Imagine you are researching bird species in Switzerland and you want to identify the variables driving the patterns of species richness.

### You will have these aims:

#### 1. Load additional bird species into the data.  
    1. Merge the environmental and bird variables and summarize your environmental data set and explain what you see.
    2. Are there any correlations or missing data? 
    3. If so, try to fill the missing values if possible, otherwise discard the rows. Find the most and least common bird species in the data.  
<br>

#### 2. Calculate Species richness  
    1. How many species do we have in total, what does this tell us about the species composition?  
    2. Calculate species richness per grid cell (add a new column with the variable species richness).  
<br>

#### 3. Build a model to explain species richness by applying a GLM with either gaussian, binomial of poisson family and the right link-function  
    1. Explain your choice of model family, report your model output (Coefficients, p values, model-fit) and give an indication of which variables lead to an increase or decrease of  species richness.   
    2. Report the estimated species richness, for 2000m elevation, 10°C temp, 600mm precipitation, 10 for slope, 20 forest.edge, 50 forest, 55 grassland, 2300m distance to water, 4 buildings.  
    3. Is there something that could be improved, is this model biological meaningful?  
<br>

#### 4. PCAs can not only inform you on correlations within data, but can be powerful in creating "independent" variables
    1. Run a PCA on the variables, extract the axes values for all axes you need to explain more than 50% of the variance, and then use these axes in a GLM with species richness.  
    2. How could you interpret the results of this model?  
    3. compared to the GLM with all the environmental variables? Elaborate on the difficulties of this approach. 
<br>

```{r 1 - example code}
# load necessary libraries
library (ade4)
library(factoextra)

# Integrate the additional bird species information to the existing data

files <- list.files("data/bird_swiss/", pattern = "*.txt") ### load all filenames in the bird_swiss folder
birdnames <- gsub(".txt", "", files ) ###remove the .TXT, gives us the cleaned names we need
bird_list <- list() ### create empty list to save the loop output

# Write a for loop reading in the presence absence matrices that we have provided and add them to the list
# the following pseudocode will help you
# for(i in ){
#   
#   # read in presence absence data for species
#   
#   pa <- read.table()
#   
#   # rename the columns after the bird species
#   
#   # save the df in the list
# 
# }

```

```{r 2 }

```

```{r 3 }

```

```{r 4 }
#you can use the factoextra package here, it has helper functions to extract PCA axis values 
#remember: the help function will provide more information if you are stuck with a package

```

